Article title

### 2017-07-24 paper-webserver.tex
ADD: more details

The surveillance of bacterial pathogens is a routine function of public-health agencies.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, culture-dependent subtyping techniques are becoming displaced by culture-independent methods based on whole-genome sequencing (WGS).
Secondary analyses, the study of relations and historical relevance of genomes, are now possible on large datasets provided by WGS \cite{ronholm2016navigating}.
Advances in WGS technologies have also enabled the monitoring and sharing of outbreak genomes in real-time \cite{ronholm2016navigating}.

Subtyping and other analyses software often require command-line or programming experience.
To make such \textit{in-silico} techniques user-friendly for front-line clinicians and laboratories, a variety of web services have been developed in Denmark \cite{joensen2014real}, Taiwan \cite{liu2016construction}, and in collaborations \cite{hasman2015detection}.
While useful, arguably \cite{fricke2014bacterial}, the lack of bioinformatic standards and common infrastructure technologies have made it difficult to fully leverage the big-data results from WGS.

Previously we developed Superphy \cite{whiteside2016superphy}, an online platform targeting \textit{E. coli}, which determines Shiga toxin subtypes, Antimicrobial Resistance (AMR) genes using the Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive}, and provides statistical significance tests between subgroups in associated metadata.
Here we present Spfy, a complete rewrite of Superphy integrating our Panseq package \cite{laing2010pan} for pan-genomic analyses of accessory regions, long-term storage of results in a graph database, and containerization of services using Docker, a modern approach found in many web servers and proposed to mitigate the complexities of bioinformatic pipelines \cite{di2015impact}.
Furthermore, Spfy takes a modular approach to tool integration and distributes analyses over task queues, thereby allowing modules to run in parallel.
All services are publicly available at \url{https://lfz.corefacility.ca/superphy/spfy/}.

>>>>

% costs of bacterial infections
As food-sourcing is globalized, consumers are at a greater risk for bacterial infections in an environment that is more complicated to track than ever before \cite{kozak2013foodborne}.
\textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common bacetrial infections \cite{kozak2013foodborne} originating from produce.
Foodborne illnesses translate into actual costs: for example, in Canada alone an estimated 4000 hospitalizations and 105 deaths per year are thought to be the result of various pathogens \cite{thomas2015estimates}.
Futhermore, the associated economic costs have been known for decades \cite{todd1989costs}.

% surveillance
The surveillance of bacterial pathogens is a routine function of public-health agencies in an effort to lower the burden caused by human illness.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, culture-dependent subtyping techniques are lately coupled with culture-independent methods based on whole-genome sequencing (WGS), to monitor known pathogens and detect emerging threats.

% traditional reference lab techniques
Subtyping encompasses various methods to designate particular strains of bacterial isolates.
In order to monitor outbreaks, platforms must rapidly subtype collected strains and communicate such information to other agencies \cite{ronholm2016navigating}. 
Traditional reference laboratory techniques handled subtyping through methods such as serotyping, detecting the presence of specific cell surface antigens as tested by antibody reaction, pulsed-field gel electrophoresis (PFGE), to identify known virulence factors (VFs) or antimicrobial resistance genes (AMR genes), or Sanger sequencing, to determine multi-locus sequence types (MLSTs) \cite{ronholm2016navigating}.
Using PFGE, outbreaks can be tracked by comparing band patterns of bacterial isolates and is routinely performed in public health agencies \cite{kozak2013foodborne}. 
The original computational methods to leverage WGS focused on creating \textit{in-silico} approaches to return subtyping results matching those of wet-lab techniques.
Software such as VirulenceFinder \cite{joensen2014real} compares a pre-collected set of reference sequences encompassing known virulence factor gene sequences against a WGS result.
Similar approaches have been applied to predict AMR genes \cite{mcarthur2013comprehensive}, O-type, and H-type \cite{whiteside2016superphy}.

% potential of WGS
WGS has the potential to expand on traditional methods, due to the inclusion of the entire genome in WGS-based analyses; in effect, this lends WGS-based methods better resolution and context in evaluating the spread of outbreaks \cite{ronholm2016navigating}.
As an example, instead of examining selected single-nucleotide polymorphisms (SNPs) in a subset of housekeeping genes, as is the case for laboratory-based MLST, newer methods such as core genome MLST (cgMLST) attempt to cover every common gene.
Secondary analyses, the study of relations and historical relevance of genomes, are now possible on large datasets provided by WGS \cite{ronholm2016navigating}.
Advances in WGS technologies have also enabled the monitoring and sharing of outbreak genomes in real-time \cite{ronholm2016navigating}.

% Drawbacks - genetics
However, these WGS-base methods still examine the underlying genetics, not the expressed phenotype.
The presence of a particular gene sequence does not necessitate expression.
Correspondingly, a greater number of target sequences are identified which complicates risk assessments \cite{ronholm2016navigating} and laboratory based phenotypic tests are still relatively inexpensive \cite(CITE ROUTINE USE - KOSER).
The effectiveness of WGS-based methods is also reliant on the accuracy and breadth of reference datasets.

% drawbacks - big data
Traditionaly, bioinformatics software was developed \textit{ad hoc}, without employing software engineering principles \cite(TRENDS - BREVEREN).
Such approaches were often script-based, with custom data formats, and only suitable for small collections of data \cite(TRENDS - BREVEREN).
As WGS-technology expanded the availability of data, systems engineering principles became important as code failure could quickly disrupt entire bioinformatics pipelines \cite(BIOLOGICAL DATA SCIENCES - SCHATZ).
The field also lags behind other big data industries: bioinformatics software typically struggles to perform computations in real-time, stores data in inflexible databases, doesn't support live visualizations \cite(TRENDS - BREVEREN), and lacks a common application programming interface (API) linking resources \cite(REVIEW - SWAMINATHAN).

% Solution via pre-computed analyses for routine tasks
Efforts such as the Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} attempt to alleviate data concerns by curating a set of known AMR genes.
In addition, agencies such as the Food and Drug Administration (FDA), which hosts GenomeTrakr \cite{allard2016practical}, the National Center for Biotechnology Information (NCBI), which hosts GenBank \cite{doi:10.1093/nar/gks1195}, and the Center for Disease Control and Prevention (CDC), which hosts PulseNet \cite(PULSENET), provide services to share genome sequences and associated metadata either openly or amongst member institutions.
Recent efforts have been focused on providing modern tools for quality control of reference genomes \cite{parks2015checkm}, particularly amongst human-genome projects \cite{lee2017ngscheckmate}.
Panseq \cite{laing2010pan}, Roary \cite{page2015roary}, and chewBBACA \url{https://github.com/mickaelsilva/chewBBACA} are available to compute core and accessory regions and identify SNPs across a large collection of reference genomes.
Taken together, open-access initiatives and shared tooling aim to alleviate the drawbacks of WGS-based methods by pre-computing sufficient reference data.

% solution via modern big data approaches
For genotypic, WGS-based approaches to be truly compliment laboratory-based phenotypic tests, analyses software must be faster or demonstrate a reasonable link between their genotype and phenotype \cite(CITE ROUTINE USE - KOSER).
Antimicrobial resistance testing, which is a major component of public health monitoring efforts, typically takes 18-24 hours in a reference laboratory \cite(CITE ROUTINE USE - KOSER).
Results should also be stored in a tiered structure, with rapid access available for retrieval and visualization while maintaining long-term data storage \cite(BIOLOGICAL DATA SCIENCES - SCHATZ).
Such an approach would enable developers to address the current limitations \cite(TRENDS - BREVEREN) of bioinformatics developers.

% why Superphy
Subtyping and reference data generation often require command-line or programming experience.
To make such \textit{in-silico} techniques user-friendly for front-line clinicians and laboratories, a variety of web services have been developed in Denmark \cite{joensen2014real}, Taiwan \cite{liu2016construction}, and in collaborations \cite{hasman2015detection}.
While useful, arguably \cite{fricke2014bacterial}, the lack of bioinformatic standards and common infrastructure technologies have made it difficult to fully leverage the big-data results from WGS.

% Superphy
Previously we developed Superphy \cite{whiteside2016superphy}, an online platform targeting \textit{E. coli}, which determines Shiga toxin subtypes, AMR genes using CARD, and provides statistical significance tests between subgroups in associated metadata and pan-genome regions.
However, we struggled to provide real-time computing of results and to accomdate an ever-growing reference database.
Furthermore, due to the nature of the various dependencies involved, the platform was difficult for other researchers to independently replicate.

% this study
Here we present Spfy, a complete rewrite of Superphy which expands on supported features to also provide long-term storage of results in a graph database, one of the emerging \cite(TRENDS - BREVEREN) database types being used for biological data.
We also containerize of services using Docker, a modern approach found in many web servers and proposed to mitigate the complexities of bioinformatic pipelines \cite{di2015impact}.
Docker integration ensures that nested dependency resolution, which typically must be manually installed \cite{doi:10.1093/bioinformatics/btu153,laing2010pan,inouye2014srst2}, are handled automatically.
Furthermore, Spfy takes a modular approach to tool integration and distrib,utes analyses over task queues, thereby allowing modules to run in parallel and addresses code failures.
These changes enable us to provide real-time genome analysis and database storage of all results for upwards of 5000 genomes.
All services are publicly available at \url{https://lfz.corefacility.ca/superphy/spfy/}.
